---
title: "BME4550_Fall2018_Assignment2"
author: "Monika Grabowska"
date: "September 11, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
```

## 5.2.4 Exercises

###1. Find all flights that 1) Had an arrival delay of two or more hours, 2) Flew to Houston (IAH or HOU), 3) Were operated by United, American, or Delta, 4) Departed in summer (July, August, and September), 5) Arrived more than two hours late, but didn’t leave late, 6) Were delayed by at least an hour, but made up over 30 minutes in flight, 7) Departed between midnight and 6am (inclusive)

###2. Another useful dplyr filtering helper is `between()`. What does it do? Can you use it to simplify the code needed to answer the previous challenges?

###3. How many flights have a missing `dep_time`? What other variables are missing? What might these rows represent?

###4. Why is `NA ^ 0` not missing? Why is `NA | TRUE` not missing? Why is `FALSE & NA` not missing? Can you figure out the general rule? (`NA * 0` is a tricky counterexample!)

## 5.3.1 Exercises

###1. How could you use `arrange()` to sort all missing values to the start? (Hint: use `is.na()`).

###2. Sort `flights` to find the most delayed flights. Find the flights that left earliest.

###3. Sort `flights` to find the fastest flights.

###4. Which flights travelled the longest? Which travelled the shortest?

## 5.4.1 Exercises

###1. Brainstorm as many ways as possible to select `dep_time`, `dep_delay`, `arr_time`, and `arr_delay` from `flights`.

###2. What happens if you include the name of a variable multiple times in a `select()` call?

###3. What does the `one_of()` function do? Why might it be helpful in conjunction with this vector?

```
vars <- c("year", "month", "day", "dep_delay", "arr_delay")
```

###4. Does the result of running the following code surprise you? How do the select helpers deal with case by default? How can you change that default?

```
select(flights, contains("TIME"))
```

## 5.5.2 Exercises

###1. Currently `dep_time` and `sched_dep_time` are convenient to look at, but hard to compute with because they’re not really continuous numbers. Convert them to a more convenient representation of number of minutes since midnight.

###2. Compare `air_time` with a`rr_time - dep_time`. What do you expect to see? What do you see? What do you need to do to fix it?

###3. Compare `dep_time`, `sched_dep_time`, and `dep_delay`. How would you expect those three numbers to be related?

###4. Find the 10 most delayed flights using a ranking function. How do you want to handle ties? Carefully read the documentation for `min_rank()`.

###5. What does `1:3 + 1:10` return? Why?

###6. What trigonometric functions does R provide?

## 5.6.7 Exercises

###1. Brainstorm at least 5 different ways to assess the typical delay characteristics of a group of flights. Consider the following scenarios: 1) A flight is 15 minutes early 50% of the time, and 15 minutes late 50% of the time.2) A flight is always 10 minutes late. 3) A flight is 30 minutes early 50% of the time, and 30 minutes late 50% of the time. 4) 99% of the time a flight is on time. 1% of the time it’s 2 hours late. Which is more important: arrival delay or departure delay?

###2. Come up with another approach that will give you the same output as `not_cancelled %>% count(dest)` and `not_cancelled %>% count(tailnum, wt = distance)` (without using `count()`).

###3. Our definition of cancelled flights (`is.na(dep_delay) | is.na(arr_delay)`) is slightly suboptimal. Why? Which is the most important column?

###4. Look at the number of cancelled flights per day. Is there a pattern? Is the proportion of cancelled flights related to the average delay?

###5. Which carrier has the worst delays? Challenge: can you disentangle the effects of bad airports vs. bad carriers? Why/why not? (Hint: think about `flights %>% group_by(carrier, dest) %>% summarise(n())`)

###6. What does the `sort` argument to `count()` do. When might you use it?

## 5.7.1 Exercises

###1. Refer back to the lists of useful mutate and filtering functions. Describe how each operation changes when you combine it with grouping.

###2. Which plane (`tailnum`) has the worst on-time record?

###3. What time of day should you fly if you want to avoid delays as much as possible?

###4. For each destination, compute the total minutes of delay. For each flight, compute the proportion of the total delay for its destination.

###5. Delays are typically temporally correlated: even once the problem that caused the initial delay has been resolved, later flights are delayed to allow earlier flights to leave. Using `lag()`, explore how the delay of a flight is related to the delay of the immediately preceding flight.

###6. Look at each destination. Can you find flights that are suspiciously fast? (i.e. flights that represent a potential data entry error). Compute the air time a flight relative to the shortest flight to that destination. Which flights were most delayed in the air?

###7. Find all destinations that are flown by at least two carriers. Use that information to rank the carriers.

###8. For each plane, count the number of flights before the first delay of greater than 1 hour.

## 7.3.4 Exercises

###1. Explore the distribution of each of the `x`, `y`, and `z` variables in `diamonds`. What do you learn? Think about a diamond and how you might decide which dimension is the length, width, and depth.

###2. Explore the distribution of `price`. Do you discover anything unusual or surprising? (Hint: Carefully think about the `binwidth` and make sure you try a wide range of values.)

###3. How many diamonds are 0.99 carat? How many are 1 carat? What do you think is the cause of the difference?

###4. Compare and contrast `coord_cartesian()` vs `xlim()` or `ylim()` when zooming in on a histogram. What happens if you leave `binwidth` unset? What happens if you try and zoom so only half a bar shows?

## 7.4.1 Exercises

###1. What happens to missing values in a histogram? What happens to missing values in a bar chart? Why is there a difference?

###2. What does `na.rm = TRUE` do in `mean()` and `sum()`?

## 7.5.1.1 Exercises

###1. Use what you’ve learned to improve the visualisation of the departure times of cancelled vs. non-cancelled flights.

###2. What variable in the diamonds dataset is most important for predicting the price of a diamond? How is that variable correlated with cut? Why does the combination of those two relationships lead to lower quality diamonds being more expensive?

###3. Install the ggstance package, and create a horizontal boxplot. How does this compare to using `coord_flip()`?

###4. One problem with boxplots is that they were developed in an era of much smaller datasets and tend to display a prohibitively large number of “outlying values”. One approach to remedy this problem is the letter value plot. Install the lvplot package, and try using `geom_lv()` to display the distribution of price vs cut. What do you learn? How do you interpret the plots?

###5. Compare and contrast `geom_violin()` with a facetted `geom_histogram()`, or a coloured `geom_freqpoly()`. What are the pros and cons of each method? 

###6. If you have a small dataset, it’s sometimes useful to use `geom_jitter()` to see the relationship between a continuous and categorical variable. The ggbeeswarm package provides a number of methods similar to `geom_jitter()`. List them and briefly describe what each one does.

## 7.5.2.1 Exercises

###1. How could you rescale the count dataset above to more clearly show the distribution of cut within colour, or colour within cut?

###2. Use `geom_tile()` together with dplyr to explore how average flight delays vary by destination and month of year. What makes the plot difficult to read? How could you improve it?

###3. Why is it slightly better to use `aes(x = color, y = cut)` rather than `aes(x = cut, y = color)` in the example above?

## 7.5.3.1 Exercises

###1. Instead of summarising the conditional distribution with a boxplot, you could use a frequency polygon. What do you need to consider when using `cut_width()` vs `cut_number()`? How does that impact a visualisation of the 2d distribution of `carat` and `price`?

###2. Visualise the distribution of carat, partitioned by price.

###3. How does the price distribution of very large diamonds compare to small diamonds? Is it as you expect, or does it surprise you?

###4. Combine two of the techniques you’ve learned to visualise the combined distribution of cut, carat, and price.

###5. Two dimensional plots reveal outliers that are not visible in one dimensional plots. For example, some points in the plot below have an unusual combination of `x` and `y` values, which makes the points outliers even though their `x` and `y` values appear normal when examined separately. Why is a scatterplot a better display than a binned plot for this case?

```{r}
ggplot(data = diamonds) +
  geom_point(mapping = aes(x = x, y = y)) +
  coord_cartesian(xlim = c(4, 11), ylim = c(4, 11))
```
